Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-07-16 sports_analytics.Rmd
jqas special issues; #11

Every sport is different, with its own set of rules, strategies, methods of data collection, number of players, and the magnitude of the role of chance. 
At the same time, many sports are similar, either because one evolved from the other, or the structure of the game shares certain attributes. 
Sports that are closely related anthropologically (NEED EXAMPLE) may or may not share common applications of analytical methods. 
Conversely, with just a few small tweaks, analytical metrics might work just as well across sports that are unrelated and quite different (e.g., an Elo rating could be equally valid for chess players and ice hockey teams).

@@

In many sports, the first step towards an analytical understanding of the game is the estimation of the expected value of a game state at any given point in a sporting contest.

Mathematically, we define $X$ to be a random variable indicating the the number of points (or runs) that a team will score over some determined amount of time (e.g. remainder of game, quarter, period, or inning). 
Let $s \in S$ be a tuple that encodes the *state* of a game. 
Then our task is to estimate:
$$
  \mathbb{E}[X | s] = \sum_{X \geq 0} \Pr[X | s] \cdot X\,,
$$
for any state $s \in S$, where $\Pr[X | s]$ is the probability of scoring $X$ points given that the game is in state $s$ and $S$ is the set of all possible states.

@@

Note that Miami starts the play with an EPV of approximately 1.0 points.
This indicates their implied average points per possession. 
Chalmers's shot is worth 3 points, so the EPV of 1.276 points implies that the model estimate of the probability of Chalmers making this shot is 42.5\%. 
A breakthrough in this work is that this estimate is conditional on the location of the other 9 players on the court!

@@

```{r chess, fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package. At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
library(chess)
# Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
game() |> plot()

library(chessR)
us <- c("fatgreggy")
greg <- get_raw_chessdotcom(
  usernames = us,
  year_month = 202206
)

m <- extract_moves_as_game(greg[6, ])
# plot_moves(m, interactive = FALSE)
plot(m)
```

@@

# Notable competitions and public events

@@

- Kaggle March Madness [@lopez2015building]
- Big Data Bowl
- Big Data Cup
- SABR Case competition

>>>>

Every sport is different, with its own set of rules, strategies, methods of data collection, number of players, and the magnitude of the role of chance. 
At the same time, many sports are similar, either because one evolved from the other, or the structure of the game shares certain attributes. 
Sports that are closely related anthropologically may or may not share common applications of analytical methods. 
For example, although belonging to the same bat-and-ball family, baseball and cricket differ in strategies such as batting order or sacrifice plays.
Conversely, with just a few small tweaks, analytical metrics might work just as well across sports that are unrelated and quite different.
For instance, an Elo rating could be equally valid for chess players and ice hockey teams.

@@

In many sports, the first step towards an analytical understanding of the game is the estimation of the expected value of a game state at any given point in a sporting contest. 
Mathematically, we define $X$ to be a random variable indicating the the number of points (or runs) that a team will score over some determined amount of time (e.g. remainder of game, quarter, period, or inning). 
Let $s \in S$ be a tuple that encodes the *state* of a game. 
Then our task is to estimate:
$$
  \mathbb{E}[X | s] = \sum_{X \geq 0} \Pr[X | s] \cdot X\,,
$$
for any state $s \in S$, where $\Pr[X | s]$ is the probability of scoring $X$ points given that the game is in state $s$ and $S$ is the set of all possible states.

@@

Note that Miami starts the play with an EPV of approximately 1.0 points.
This indicates their implied average points per possession. 
Chalmers's shot is worth 3 points, so the EPV of 1.276 points implies that the model estimate of the probability of Chalmers making this shot is 42.5\%. 
A breakthrough in this work is that this estimate is conditional on the locations of the other 9 players on the basketball court!

@@

```{r chess, fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package. At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
# library(chess)
# # Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
# game() |> plot()
# 
# library(chessR)
# us <- c("fatgreggy")
# greg <- get_raw_chessdotcom(
#   usernames = us,
#   year_month = 202206
# )
# 
# m <- extract_moves_as_game(greg[6, ])
# # plot_moves(m, interactive = FALSE)
# plot(m)
```

@@

# Opportunities

@@

- Kaggle March Madness [@lopez2015building] (jqas special issue: https://www.degruyter.com/journal/key/jqas/11/1/html)
- Big Data Bowl (special issue https://www.degruyter.com/journal/key/jqas/16/2/html)
- Big Data Cup
- SABR Case competition
